---
title: "Lab4"
format: 
  html:
    self-contained: true
editor: visual
code-fold: true
theme: mint
execute:
  warning: false
  echo: true
  message: false 
---

# 2.

```{r set_up}
library(tidyverse)
library(forcats)
avocados <- read_csv( here::here("supporting_artifacts", "ref.work", "lab4", "avocado.csv"))

```

-   The dataset contains sales data for hass avocados sold from 2018 based on bar codes scanned in stores. The dataset contains data from 2013 - 2017.

### 2.2

```{r data_clean}
#| include: false
avocado_clean <- avocados |> 
  rename(
    "small" = "4046", "large" = "4225", "xlarge" = "4770",
    "Total.Volume" = "Total Volume"
    ) 

avocado_clean <- avocado_clean |> 
  separate(Date, sep = "-", into = c("year", "month","day")) |> 
  mutate(year = as.numeric(year),
         month = as.numeric(month))
```

## 3.

```{r small_avo}
regions <- 
  as_tibble_col(c("Plains", "SouthCentral", "Southeast", 
                  "West", "Northeast", "Midsouth", 
                  "WestTexNewMexico", "GreatLakes" ), 
                column_name = "MajorRegions")

semi_join(avocado_clean, regions, 
          c("region" = "MajorRegions")) |> 
  filter(year == 2017) |> 
  slice_max(small) |> 
  select(region, small)
```

-   the south central sold the most avocados

### 3.2

```{r monthly_sales}

avocado_clean |> 
  select(month, Total.Volume) |> 
  group_by(month) |> 
  summarise(mean_volume = mean(Total.Volume)) |> 
  slice_max(mean_volume)
```

-   February has the highest volume of sales.

### 3.3

```{r state_comparison}
states <- as_tibble_col(
  c("California", "SouthCarolina","TotalUS"), 
  column_name = "state"
  )

avocado_metro <- avocado_clean |> 
  anti_join(states,
            by = c("region" = "state")) |> 
   anti_join(regions, 
            by = c("region" = "MajorRegions"))
 
avocado_metro <- avocado_metro |> 
  select(region, Total.Volume) |> 
  group_by(region) |> 
  summarise(mean_total = mean(Total.Volume), 
            .keep_all = TRUE) |> 
  slice_max(mean_total, n = 5)
top_5 <- as_tibble_col(avocado_metro$region, 
                       column_name = "city")
avocado_metro <- 
  semi_join(avocado_clean, top_5, 
            by = c("region" = "city")) |> 
  select(region, Total.Volume)

ggplot(data = avocado_metro,
       mapping = aes(x = Total.Volume, 
                     y =fct_reorder(region, Total.Volume, 
                                    .desc = TRUE),
                     color = region)) +
  geom_boxplot(outlier.shape = NA, show.legend = FALSE) +
  xlab("Avocados Sold") +
  ylab("Metropolitin Area")
```

## 4. Reshaping

```{r OrganicVsConventional}
cali <- as_tibble_col(c("LosAngeles", "SanDiego", "Sacramento", "SanFrancisco"), 
                      column_name = "city")

avocado_cali <- 
  semi_join(avocado_clean, cali, c("region" = "city"))

avocado_cali |> 
  select(region, type, Total.Volume, AveragePrice) |> 
  group_by(region, type) |>
  summarize(MeanPrice = mean(AveragePrice),
            MedianPrice = median(AveragePrice)) |> 
  pivot_wider(names_from = type, 
              values_from = c(MeanPrice, MedianPrice)) |> 
  mutate(
    MeanPriceDiff = MeanPrice_organic - MeanPrice_conventional,
    MedianPriceDiff = MedianPrice_organic - 
                      MedianPrice_conventional
    ) |> 
  arrange(desc(MeanPriceDiff)) |> 
  select(region, MeanPriceDiff, MedianPriceDiff)
```

-   On average the average and median largest difference between the price of organic and conventional avocados is in San Francisco. The largest average and median percent difference in the price is in San Diego but median percent difference is very close to San Francisco.

### 4.2

```{r california_graph}
avocado_cali <- avocado_cali |> 
  select(region, small, large, xlarge, Total.Volume, type) |> 
  group_by(region, type) |> 
  mutate(small = mean(small),
         large = mean(large),
         xlarge = mean(xlarge),
         Total.Volume = mean(Total.Volume)) |> 
  distinct(region, .keep_all = TRUE)

#pivot_longer
avocado_cali <- avocado_cali |> 
  pivot_longer(
    c(small, large, xlarge),
    names_to = "size", values_to = "proportion"
    ) |> 
  select(region, proportion, size, type) |> 
  group_by(region) |> 
  rename(`Avocado Size`= "size")

ggplot(data = avocado_cali) +
  geom_col(
    mapping = aes(x = region, y = proportion,  
                  fill = `Avocado Size`), 
    position = "fill"
    ) +
  ylab("Proportion of Mean Avocados Sold") +
  xlab("Region of CA") +
  facet_wrap(~type, nrow = 1) +
  scale_x_discrete(guide = guide_axis(n.dodge=2))
  
# I used https://datavizpyr.com/how-to-dodge-overlapping-text-on-x-axis-labels-in-ggplot2/ to help with formatting the xlabels
```
